Theories of deliberated choice

### 2022-08-15 simple.tex
Phrasing and bib

\begin{document}
\title{Theories of deliberated judgment}
\author{Olivier Cailloux}
\affil{Université Paris-Dauphine, PSL Research University, CNRS, LAMSADE, 75016 PARIS, FRANCE\\
	\href{mailto:olivier.cailloux@dauphine.fr}{olivier.cailloux@dauphine.fr}
}
\makeatletter
	\hypersetup{
		pdfsubject={Epistemology},
		pdfkeywords={Decision aiding, Decision making, Argumentation}
	}
\makeatother
\maketitle

@@

\section{Introduction} 
Observing subjects performing acts of choice is the basis of much work in econometrics. Such acts have been used, among others, to give an observable basis to the concept of preference \citep{samuelson_foundations_1983}. %Acts of choice are considered here in the broad sense of choice between courses of actions, encompassing the simple setting of choosing between various elementary consumption bundles as well as settings involving more abstract acts such as choosing a strategy in a game-theoretical setting.

@@

While numerous articles have discussed the concept of deliberation in the last decades, it remains unsettled, to the best of my knowledge, how to observe a deliberated judgment, or, correspondingly, how to define this concept in a way that makes it amenable to empirical study, sufficiently precisely that any disagreement about someone’s deliberated judgment in a given context could be solved, at least in principle, by an empirical experiment. %This article is interested in defining the concept of deliberated judgementof a subject, as opposed to the classical one, that will henceforth be called her natural judgment, that is, the way that she chooses between acts when she does not face any arguments, i.e.,

While the concept of deliberation in the deliberative democracy literature usually refers to multiple individuals debating, this article considers deliberation as related to a single individual at a time: an individual deliberates (in the sense of: carefully weights) the stance she considers most appropriate while being confronted to arguments. The deliberated judgment of an individual is thus defined with no requirement that the arguments be uttered by peers in a debate; they can be given by an automata, for example. In this sense, deliberated judgments come close to what Rawls calls reflective equilibrium (referring to an idea of Goodman), a concept that has much inspired the present proposal.

To define deliberated judgments, one needs to start with the set of arguments under consideration, and the way the subject gets to know them, called hereafter the “exposure protocol”. As a result, we need always talk about the deliberated judgment \emph{given some exposure protocol}. The notion of argument invoked here is an extremely large one, including texts, images, sounds, experiences (a hiking trip), and so on: anything that can be transmitted to a subject qualifies as a possible argument. Such an “argument” need not even have the logical structure of what is considered a proper argument in argumentation theory, which permits to avoid having to take position on what is a correct argument, and permits to study wide enough conceptions of deliberated judgment, including those involving no paternalism: anything that influences the judgment of a subject may a priori be considered worth including in at least some conceptions of deliberated judgment, thus including whatever some could consider “bad reasoning” or “bad reasons”.

The dependence of the deliberated judgment concept on the exposure protocol yields a first qualification about the concept of deliberated judgments defined here when one desires to use it normatively: it is only as reasonable to call the resulting judgments “deliberate” in a normative sense as the exposure protocol is normatively relevant to deliberation. A sufficiently restricted, or even purposefully chosen, set of arguments may yield judgments that are only deliberate in the technical sense used here, but that pertain more to brain-washing than to reflective judgments. While it is indeed a long-term goal of this research program to yield a normatively relevant conception of preference (appropriate, for example, for recommendation), it is not a claim of this article that anything that corresponds to deliberated judgments as defined here qualify as being appropriate for recommendation (or for any other normative use): much depends on the chosen exposure protocol. The goal of this article is, more modestly, one of a conceptual and practical clarification. It proposes a sharp distinction between the empirical content and the normative content of the concept of deliberated judgments: the definition proposed here transfers the full normative weight on the choice of the exposure protocol; and leaves the rest of the disagreements to empirical settling. This is methodologically akin, mutatis mutandis, to the axiomatic study of justice such as done in social choice, only replacing mathematical deduction with empirical claims. In social choice, the choice of axioms bears the full normative weight; once axioms are chosen, it is a matter of deduction which social choice rules (if any) correspond to that choice. In our case, once an exposure protocol is chosen, it is a pure empirical matter which deliberated judgments result.
In supplement to a possibly helpful conceptual clarification, such a sharp separation may reveal practically useful for separation of concern as it permits to study deliberated judgments empirically under various exposure protocols independently of possible persistent disagreements about which norms are more appropriate for deliberation in which circumstances.

The deliberated judgment of a subject is said to depend on the exposure protocol, that is, on the set of arguments and the way they are submitted to the subject, and not only on the set of arguments, because the very way that they are transmitted to the individual may affect her judgment differently. For example, reading an image description and showing an image to a subject may impress them differently.

The exposure protocol defines a set of argument and, importantly, how arguments are sent to the subject and her subsequent choice observed. There is, as need be, no requirement on the internal process that occurs in the subject’s mind while she (possibly) processes the arguments, as it is not desirable to assume that the internal process could be observed. We consider subjects as black boxes and only require a capacity of observation of their choices after exposure to arguments. Depending on the exposure protocol and the subject, it is of course possible for the arguments to never affect any choice of the subject, a conclusion that may be interesting in its own right.

@@

Studying judgments resulting from exposure to arguments includes a difficulty that is not present with natural choice experiments: exposing an individual to arguments may change the individual’s stance, an effect that cannot be erased to submit the individual (considered “clean” again) to another sequence of arguments. The observer thus cannot in general assume that multiple unrelated sequences of arguments can be tested on a given individual. This difficulty will be worked around by considering general theories of deliberated judgments, that apply to sets of individuals. This is methodologically similar to defining theories of how objects break. One can test only one way of breaking a given (unrepairable) object, so such theories must apply to sets of objects if one wants to be able to test multiple ways of falsifying the theories.

@@

While $I$ and $\allargs$ are fundamental sets whose elements are considered elementary, a few notations and concepts are required to define ${\allleadsto}$.
Throughout the article, $\N$ includes zero. %and $\N^* = \N \setminus \set{0}$. 
Given $j, l \in \N$, the notation $\intvl{j, l}$ represents $\set{k \in \N \suchthat j ≤ k ≤ l}$ (thus $\forall j \in \N: \intvl{j, 0} = \emptyset$).
Define $\allhist = \bigcup_{k \in \N} \allargs^{\intvl{1, k}}$ as the finite sequences of arguments, including the empty sequence. 
A generic element of $\allhist$ (a sequence of arguments) is denoted by $\hist$ and called an argumentative path, or simply a path when this raises no ambiguity. Its length is denoted by $\card{\hist} \in \N$.
Thus, given an argumentative path $\hist \in \allhist$, $\forall k \in \intvl{1, \card{\hist}}: \hist_k \in \allargs$, and $\hist = \emptyset ⇔ \card{\hist} = 0$.

@@

\subsection{Deliberated choice}
The deliberated choice of $i$ given the exposure protocol $(I, \allargs, {\allleadsto})$ is denoted $P_i \subset P$.
Its formal defnition requires the notion of decisive argument.

@@

\subsection{An example and a discussion about observability}
The following toy example is inspired by citet[p.~59]{Railton}\commentOC{TODO}, who uses this example to support his claim that beliefs may influence decisions. Here is his story.
\begin{quote}
An experiment in cognitive psychology reveals that when subjects observe two individuals taking a test, one of whom answers a high percentage of the early questions correctly but then falters, while the other does
poorly initially but then answers a high percentage of the later questions
correctly, the first is usually viewed as more able, even though in the end
each answers the same number of questions correctly in all. If the subjects had been asked which test-taker they would prefer as a math tutor,
they likely would have answered “The first,” and perhaps they would also
have been willing to pay a premium to secure his services. But if they
were to become convinced of the influence of sheer order on their evaluations of relative ability, they presumably would no longer want their
initial preference for the first test-taker to influence their choices – or
willingness to pay – in this way.
\end{quote}
\begin{example}[Math tutor]
	\label{ex:tutor}
	Let $I$ be a set of individuals desiring to evaluate two candidates $\set{m_1, m_2}$ for math tutoring.
	Let ${\leadsto}(i, a_1)$ represent $i$ being shown a test of candidate $m_1$ wherein $m_1$ answers 70\% of the first half of questions correctly but then only 40\% of the other half correctly.
	Let ${\leadsto}(i, a_2)$ represent $i$ being shown a test of candidate $m_2$ wherein $m_2$ answers 45\% of the first half of questions correctly, then 75\% of the other half correctly.
%	Let ${\leadsto}(i, a_e)$ represent $i$ being given an explanation about the psychological phenomenon described in the quote above.
	Let ${\leadsto}(i, a_e)$ represent $i$ being given the following argument. “When tested, $m_1$ answers 70\% of the first half of questions correctly but then only 40\% of the other half correctly; whereas $m_2$ answers 45\% of the first half of questions correctly, then 75\% of the other half correctly. It follows than $m_1$ answers on average less well than $m_2$. Observing the candidates take the tests may yield a contrary impression; this is a well-known psychological phenomenon depending on the order of the questions that the candidates answer correctly.”
	Let $\Phi = \set{\prop_1, \prop_2}$ where $\prop_1$ represent $m_1$ and $\prop_2$ represent $m_2$.

	If the prediction described in the quote is correct, and applies equally to this modified experiment, then, given $i \in I$, $i$’s deliberated choice is $\prop_2$ because $a_e$ is a decisive argument for $\prop_2$, even though $i$ might have rather chosen $\prop_1$ if given only the arguments $a_1$ and $a_2$ without $a_e$.
\end{example}

That the deliberated choice of any individual $i$ is well-defined may be doubtful, as it involves counterfactuals.
To see this, let $\alpha_{123} = (a_1, a_2, a_e)$ designate the history where the arguments are shown in the order they are described in the example, and let $\alpha_{321} = (a_3, a_2, a_1)$ designate the reverse order.
The deliberated choice of $i$ is defined so as to not depend on the (arbitrary) order of the arguments shown to $i$: for $\prop \in \iprops$, it must be that directing to $i$ the path $\alpha_{123}$ or the path $\alpha_{321}$ both lead $i$ to choose $\prop$.
However, once we have shown to $i$, say, $\alpha_{123}$, we cannot show to her $\alpha_{321}$ (unless assuming $i$ forgets arguments after a while and waiting for sufficiently long, which would create practical and theoretical difficulties).

This difficulty is resolved using falsifiable general theories, applicable to $I$ and not solely to a given $i$, so that, intuitively speaking, $\alpha_{321}$ can be tested on some $i' ≠ i$. Indeed, what is needed is a way to make a claim such as the one concluding \cref{ex:tutor} empirically testable. This is achieved in the next section thanks to theories of deliberated choice.
% Such theories contain an observable claim and a conclusion claim, so as to satisfy two properties. First, such a theory must be falsifiable in principle, meaning that it must be possible, using our observational basis (propositions of the sort $\alpha \ileadsto \phi$), to invalidate the observable claim of the theory. As a counter-example, using \cref{ex:tutor}, a theory that claims that $\forall i \in I: \alpha_{123} \ileadsto \prop_1 \lor \alpha_{321} \ileadsto \prop_1$ is not falsifiable in principle, as, whatever the real $\ileadsto$ relation, with a given $i$, one can never show that $\alpha_{123} \nileadsto \prop_1 \land \alpha_{123} \nileadsto \prop_1$. Second, such a theory must be so that if its observable claim holds, then some given alternative is in the deliberated judgment of the individuals in $I$.

\section{Theories of deliberated choice}
This section assumes that an exposure protocol $(I, \allargs, {\leadsto})$ is defined, and defines theories of deliberated choice associated with that exposure protocol. We will start with very simple theories in order to build an intuition about the conditions required on theories of deliberated choice (\cref{sec:static}), then give an example to illustrate why more complex theories are needed (\cref{sec:lichtenstein}) and finally define more general theories of deliberated choice (\cref{sec:dynamic}).

@@

In many decision problems, individuals will not be attracted by the same set of arguments and counter-arguments, even though they may finally opt for the same alternative. Imagine that some individual $i_1$ considers an argument $a_1$ for $\nprop$ relevant and some individual $i_2$ considers an argument $a_2$ for $\nprop$ relevant, but $i_1$ does not consider $a_2$ relevant and $i_2$ does not consider $a_1$ relevant (for example, because they already have some counter-arguments in mind to those arguments). Imagine further that there is some counter-argument $a'_1$ that can be given to $i_1$ to rebut $a_1$, and a different counter-argument $a'_2$ that can be given to $i_2$ to rebut $a_2$, so that they both have $\prop$ in their deliberated judgment when considering all these arguments.
In such situation, static theories as defined in \cref{sec:static} are ill suited: these theories need to present the same argument to all individuals and would thus need to define a bigger argument $a$ containing both $a'_1$ and $a'_2$. Such an argument could be too long to be convincing to $i_1$ or $i_2$, as it needlessly spends efforts countering an argument that the individual considers irrelevant to start with. This  will become an important problem if there are not only two but many a priori possible arguments $a_k$ for $\nprop$, a small subset of which is considered relevant by a given individual, each with their own counter-argument $a'_k$ for $\prop$.

%For a partial illustration about an ambitious setting, consider a decision to be taken about the relevance of some concrete action to help cope with climate change, and let $I$ be a set of citizens interested in the climate issue. The complete set of all arguments that will be considered relevant by at least one individual may well be very large. It may also be that, for each individual who considers relevant some arguments $A$ against the proposed concrete action, it is possible to give counter-arguments $A'$ that will convince the individual to opt for the proposal.
One of the settings discussed by \citet{thaler_nudge_2009} permits to illustrate this point more concretely.
Let $I$ be a set of workers interested in establishing a savings plan. 
The authors claim that workers tend to not save enough because of psychological biases. 
It is therefore tempting to conjecture that the deliberated choice of those workers is to save more, in the sense defined here, than what their natural choice (the one revealed in absence of arguments) leads to.
A difficulty related to this decision problem is that the complete set of all arguments that will be considered relevant by at least one individual for such a complex issue may well be very large. It may also be that, for each individual who considers relevant some arguments $A$ for saving not much (i.e., spending more in the short term), it is possible to give counter-arguments $A'$ that will convince the individual to opt for saving more.
But no single argument could be decisive for each $i \in I$: it could be that only tailored counter-arguments could convince individuals, given the complexity of the issue.
This illustrates that the deliberated choice of each individuals in $I$ may be to save more even though no static theory (\cref{def:static}) may achieve showing it, as it could be that no such theory is truthful.

@@

Define $\allprops = \set{\phi_P, \phi_D}$.

Define $\ar_e$ as a text arguing that the choice between $P$ and $D$ should be determined by their expected revenue, namely, $\$3.959$ for $P$ and $\$4$ for $D$.
\footnote{Here is a possible text for $\ar_e$. “Imagine you play $P$ 100 times. On average, you would win 99 times and lose once, which means gaining 99 × \$4 = \$396 and losing 1 × \$0.10 = \$0.10, hence a total net gain of \$395.90. Thus, your average net gain for the $P$ bet would be \$3.9590. [Similar text for $D$.] Therefore, we suggest you choose $D$ for a higher expected revenue.”}
Define $\ar_u$ as a text arguing that the utility of money should matter, not the revenue, and arguing that a loss hurts more than a gain. 
\footnote{Here is a possible text for $\ar_u$. “$D$ is better than $P$ in terms of expected gain, but it is in general absurd to reason in terms of expected gain. For example, consider a choice between winning $10^6$ dollars with probability $0.8$ and winning $2 × 10^6$ dollars with probability $0.4$. These two lotteries have the same expected gain, but the first one is clearly preferrable. Therefore, we recommend to rather pick $P$ as it presents a much better chance of earning money”.}
Define a counter-argument to $\ar_u$, $\ar_l$, indicating that for small monetary amounts, it is perfectly reasonable to consider utility as linear, especially given that in this experiment, multiple lotteries will be played.
\footnote{Here is a possible text for $\ar_l$. “The lottery $D$ has a higher expected revenue, and thus, a higher utility than $P$, because for small monetary amounts, it is perfectly reasonable to consider utility as linear, especially given that in this experiment, multiple lotteries will be played.”}
Define $\ar$ so as to include the contents of $\ar_e$, $\ar_u$ and $\ar_l$.
\footnote{Here is a possible text for $\ar$. [Text of $\ar_e$] “Note that one could say that the utility of money should matter, but for small monetary amounts, it is perfectly reasonable to consider utility as linear, especially given that in this experiment, multiple lotteries will be played.”}
Define $\allargs = \set{\ar_e, \ar_u, \ar_l, \ar}$.

@@

Assume that some individuals $i \in I$ do not see the point of argument $\ar_u$ and some do but are convinced by the counter-argument $\ar_l$. Accordingly, define $\ileadsto[e]$, representing the first kind of individuals, as follows. $\forall \hist \in \allhist: \hist \ileadsto[e] D ⇔ [\ar_e \in R(\hist)]$, and $\hist \ileadsto[e] P$ otherwise. Note that in particular, if $\hist$ contains $\ar$ but not $\ar_e$, $\ar \nileadsto[e] D$, as we assume such individuals are confused by the longer argument $\ar$ referring to abstract notions and therefore opt for $P$, falling back to the common psychological bias.
The second kind of individuals is represented by $\ileadsto[u]$ defined as $\forall \hist \in \allhist: \hist \ileadsto[u] D ⇔ [\ar_u \notin R(\hist) \lor \ar \in R(\hist) \lor \set{\ar_u, \ar_l} \subseteq R(\hist)]$, and $\hist \ileadsto[u] P$ otherwise.
We assume here that such individuals think about $\ar_u$ as a correct argument in favor of $P$ until shown the rebuttal $\ar_l$ (or $\ar$, which contains the same rebuttal).
Assuming finally $\allleadsto$ as equal to $\ileadsto[e]$ for individuals of the first kind and $\ileadsto[u]$ otherwise concludes the definition of the exposure protocol $(I, \allargs, {\allleadsto})$.

@@

>>>>

\begin{document}
\title{Theories of deliberated choice}
\author{Olivier Cailloux}
\affil{Université Paris-Dauphine, PSL Research University, CNRS, LAMSADE, 75016 PARIS, FRANCE\\
	\href{mailto:olivier.cailloux@dauphine.fr}{olivier.cailloux@dauphine.fr}
}
\makeatletter
	\hypersetup{
		pdfsubject={Epistemology},
		pdfkeywords={Decision aiding, Decision making, Argumentation}
	}
\makeatother
\maketitle

@@

\section{Introduction} 
Observing subjects perform acts of choice is the basis of much work in econometrics. Such acts have been used, among others, to give an observable basis to the concept of preference \citep{samuelson_foundations_1983}. %Acts of choice are considered here in the broad sense of choice between courses of actions, encompassing the simple setting of choosing between various elementary consumption bundles as well as settings involving more abstract acts such as choosing a strategy in a game-theoretical setting.

@@

While numerous articles have discussed the concept of deliberation in the last decades, it remains unsettled, to the best of my knowledge, how to observe a deliberated judgment in a systematic and reproducible fashion. There is a need to define this concept sufficiently precisely that, first, it is clear what counts as a deliberated judgment (i.e., when “sufficient” \citep{meinard_justification_2020} or “correct” deliberation has happened); and second, it is clear which part of the claim is empirical and which part stems from an unfalsifiable conception of what “deliberated” means. This permits to study deliberated judgments empirically, meaning here that disagreements about someone’s deliberated judgment in a given context can be solved, at least in principle, by an empirical experiment, provided the very conception of what constitutes a deliberated judgment is shared. This article is interested in defining the concept of deliberated judgement of a subject and show how one can study it in an emprical manner. The “classical” counterpart of deliberated judgments,, that does not involve confrontation with arguments, will henceforth be called the natural judgment. As we deal with choice acts, we will also name the concept “deliberated choice”, as opposed to “natural choice”.

While the concept of deliberation in the deliberative democracy literature usually refers to multiple individuals debating, this article considers deliberation as related to a single individual at a time: an individual deliberates (in the sense of: carefully weights) the stance she considers most appropriate while being confronted to arguments. The deliberated judgment of an individual is thus defined with no requirement that the arguments be uttered by peers in a debate; they can be given by an automata, for example. In this sense, deliberated judgments come close to what \citet{rawls_theory_1999} calls reflective equilibrium (referring to an idea of \citet{goodman_fact_1983}), a concept that has much inspired the present proposal.

To define deliberated judgments, one needs to start with the set of arguments under consideration, and the way the subject gets to know them, called hereafter the “exposure protocol”. As a result, we need always talk about the deliberated judgment \emph{given some exposure protocol}. The notion of argument invoked here is an extremely large one, including texts, images, sounds, experiences (a hiking trip), and so on: anything that can be transmitted to a subject qualifies as a possible argument. In particular, such an “argument” need not have the logical structure of what is considered a proper argument in argumentation theory. This important feature of this proposal permits to avoid taking a position on what is a correct argument, and permits to study wide enough conceptions of deliberated judgment, including those involving no paternalism \citep{cailloux_formal_2020}: anything that influences the judgment of a subject, as judged by the subject herself and not by any external standard, may a priori be considered worth including in at least some conceptions of deliberated judgment, thus including whatever some could consider as “incorrect reasoning” or “bad reasons”.

The dependence of the deliberated judgment concept on the exposure protocol yields a first qualification about the concept of deliberated judgments when one desires to use it normatively: it is only as reasonable to call the resulting judgments “deliberate” in a normative sense as the exposure protocol is normatively relevant to deliberation. A sufficiently restricted, or even purposefully chosen, set of arguments may yield judgments that are only deliberate in the technical sense used here, but that pertain more to brain-washing than to reflective judgments. While it is indeed a long-term goal of this research program to yield a normatively relevant conception of preference (appropriate, for example, for recommendation), it is not a claim of this article that anything that corresponds to deliberated judgments as defined here qualify as being appropriate for any normative use: much depends on the chosen exposure protocol. The goal of this article is one of a conceptual and practical clarification. It proposes a sharp distinction between the empirical content and the normative content of the concept of deliberated judgments: the definition proposed here transfers the full normative weight on the choice of the exposure protocol; and leaves the rest of the disagreements to empirical settling. This is methodologically akin, mutatis mutandis, to the axiomatic study of justice such as done in social choice, only complementing mathematical deduction with empirical claims. In social choice, the choice of axioms bears the full normative weight; once axioms are chosen, it is a matter of logic which social choice rules (if any) correspond to that choice. In our case, once an exposure protocol is chosen, it is an empirical matter which deliberated judgments result.
In supplement to a possibly helpful conceptual clarification, such a sharp separation may reveal practically useful for separation of concern as it permits to study deliberated judgments empirically under various exposure protocols independently of possible disagreements about which norms are more appropriate for deliberation in which circumstances.

The deliberated judgment of a subject is said to depend on the exposure protocol, that is, on the set of arguments and the way they are submitted to the subject, and not only on the set of arguments, because the very way that they are transmitted to the individual may affect her judgment differently. For example, reading an image description and showing an image to a subject may impress them differently \citep{railton_facts_2003}.

The exposure protocol defines a set of argument and, importantly, how arguments are sent to the subject and her subsequent choice observed. There is no explicit dependency on the internal process that occurs in the subject’s mind while she (possibly) processes the arguments, as it is not desirable to assume that the internal process could be observed. We consider subjects as black boxes and only require a capacity of observation of their choices after exposure to arguments. Depending on the exposure protocol and the subject, it is of course possible for the arguments to never affect any choice of the subject, a conclusion that may be interesting in its own right.

@@

Studying judgments resulting from exposure to arguments includes a difficulty that is not present with natural choice experiments: exposing an individual to arguments may change the individual’s stance, an effect that cannot be erased in order to submit the individual to an unrelated sequence of arguments. The observer thus cannot in general assume that multiple unrelated sequences of arguments can be tested on a given individual. This difficulty will be worked around by considering general theories of deliberated judgments, that apply to sets of individuals. This is methodologically similar to defining theories of how objects break. One can test only one way of breaking a given (unrepairable) object, so such theories must apply to sets of objects if one wants to be able to test multiple ways of falsifying the theories.

@@

While $I$ and $\allargs$ are fundamental sets whose elements are considered elementary, a few notations and concepts are required to define ${\allleadsto}$.
Throughout the article, $\N$ includes zero. %and $\N^* = \N \setminus \set{0}$. 
Given $j, l \in \N$, the notation $\intvl{j, l}$ represents $\set{k \in \N \suchthat j ≤ k ≤ l}$ (thus $\forall j \in \N: \intvl{j, 0} = \emptyset$).
Define $\allhist = \bigcup_{k \in \N} \allargs^{\intvl{1, k}}$ as the finite sequences of arguments, including the empty sequence. 
A generic element of $\allhist$ (a sequence of arguments) is denoted by $\hist$ and called an argumentative path, or simply a path. Its length is denoted by $\card{\hist} \in \N$.
Thus, given an argumentative path $\hist \in \allhist$, $\forall k \in \intvl{1, \card{\hist}}: \hist_k \in \allargs$, and $\hist = \emptyset ⇔ \card{\hist} = 0$.

@@

\subsection{Deliberated choice}
The deliberated choice of $i$ given the exposure protocol $(I, \allargs, {\allleadsto})$ is denoted $P_i \subset P$.
Its formal definition requires the notion of decisive argument.

@@

\subsection{An example and a discussion about observability}
The following toy example is inspired by an argument from \citet[p.~59]{railton_facts_2003}. Here is his argument.
\begin{quote}
An experiment in cognitive psychology reveals that when subjects observe two individuals taking a test, one of whom answers a high percentage of the early questions correctly but then falters, while the other does
poorly initially but then answers a high percentage of the later questions
correctly, the first is usually viewed as more able, even though in the end
each answers the same number of questions correctly in all. If the subjects had been asked which test-taker they would prefer as a math tutor,
they likely would have answered “The first,” and perhaps they would also
have been willing to pay a premium to secure his services. But if they
were to become convinced of the influence of sheer order on their evaluations of relative ability, they presumably would no longer want their
initial preference for the first test-taker to influence their choices – or
willingness to pay – in this way.
\end{quote}
\begin{example}[Math tutor]
	\label{ex:tutor}
	Let $I$ be a set of individuals desiring to evaluate two candidates $\set{m_1, m_2}$ for math tutoring.
	Let ${\leadsto}(i, a_1)$ represent $i$ being shown a test of candidate $m_1$ wherein $m_1$ answers 70\% of the first half of questions correctly but then only 40\% of the other half correctly.
	Let ${\leadsto}(i, a_2)$ represent $i$ being shown a test of candidate $m_2$ wherein $m_2$ answers 45\% of the first half of questions correctly, then 75\% of the other half correctly.
%	Let ${\leadsto}(i, a_e)$ represent $i$ being given an explanation about the psychological phenomenon described in the quote above.
	Let ${\leadsto}(i, a_e)$ represent $i$ being given the following argument. “When tested, $m_1$ answers 70\% of the first half of questions correctly but then only 40\% of the other half correctly; whereas $m_2$ answers 45\% of the first half of questions correctly, then 75\% of the other half correctly. It follows than $m_1$ answers on average less well than $m_2$. Observing the candidates take the tests may yield a contrary impression; this is a well-known psychological phenomenon depending on the order of the questions that the candidates answer correctly.”
	Let $\prop_1$ represent the choice of $m_1$ and $\prop_2$ that of $m_2$.

	If the prediction described in the quote is correct when applied to this modified experiment, then, given $i \in I$, $i$’s deliberated choice is $\prop_2$ because $a_e$ is a decisive argument for $\prop_2$, even though $i$ might have rather chosen $\prop_1$ if given only the arguments $a_1$ and $a_2$ without $a_e$.
\end{example}

That the deliberated choice of any individual $i$ is well-defined may be doubtful, as it involves counterfactuals.
To see this, let $\alpha = (a_1, a_2, a_e)$ designate the argumentative path where the arguments are shown in the order they are described in the example, and let $\alpha' = (a_e, a_2, a_1)$ designate the reverse order.
The deliberated choice concept is defined so as to not depend on the (arbitrary) order of the arguments shown to $i$: for $\prop \in P_i$, it must be that directing to $i$ the path $\alpha$ or the path $\alpha'$ both lead $i$ to choose $\prop$.
However, once we have shown $\alpha$ to $i$, we cannot show to her $\alpha'$ (unless assuming $i$ forgets arguments after a while and waiting for sufficiently long, which would create practical and theoretical difficulties).

This difficulty is resolved using falsifiable general theories, applicable to $I$ and not solely to a given $i$, so that $\alpha$ can be tested on $i$ and $\alpha'$ on some $i' ≠ i$. Indeed, what is needed is a way to make a claim such as the one concluding \cref{ex:tutor}, according to which $a_e$ is a decisive argument, empirically testable. This is achieved in the next section thanks to theories of deliberated choice.
% Such theories contain an observable claim and a conclusion claim, so as to satisfy two properties. First, such a theory must be falsifiable in principle, meaning that it must be possible, using our observational basis (propositions of the sort $\alpha \ileadsto \phi$), to invalidate the observable claim of the theory. As a counter-example, using \cref{ex:tutor}, a theory that claims that $\forall i \in I: \alpha_{123} \ileadsto \prop_1 \lor \alpha_{321} \ileadsto \prop_1$ is not falsifiable in principle, as, whatever the real $\ileadsto$ relation, with a given $i$, one can never show that $\alpha_{123} \nileadsto \prop_1 \land \alpha_{123} \nileadsto \prop_1$. Second, such a theory must be so that if its observable claim holds, then some given alternative is in the deliberated judgment of the individuals in $I$.

\section{Theories of deliberated choice}
This section assumes that an exposure protocol $(I, \allargs, {\leadsto})$ is defined, and defines theories of deliberated choice associated with that exposure protocol. We will start with very simple theories (\cref{sec:static}) in order to build an intuition about the conditions required on theories of deliberated choice, then justify the need for more complex theories (\cref{sec:lichtenstein}) and finally define more general theories of deliberated choice (\cref{sec:dynamic}).

@@

In many decision problems, individuals will not be attracted by the same set of arguments and counter-arguments, even though they may finally opt for the same alternative. Imagine that some individual $i_1$ considers an argument $a_1$ for $\nprop$ relevant and some individual $i_2$ considers an argument $a_2$ for $\nprop$ relevant, but $i_1$ does not consider $a_2$ relevant and $i_2$ does not consider $a_1$ relevant (for example, because they already have some counter-arguments in mind to those arguments). Imagine further that there is some counter-argument $a'_1$ that can be given to $i_1$ to rebut $a_1$, and a different counter-argument $a'_2$ that can be given to $i_2$ to rebut $a_2$, so that they both have $\prop$ in their deliberated judgment when considering all these arguments.
In such situation, static theories as defined in \cref{sec:static} are ill suited: these theories need to present the same argument to all individuals and would thus need to define a bigger argument $a$ containing both $a'_1$ and $a'_2$. Such an argument could be too long to be convincing to $i_1$ or $i_2$, as it needlessly spends efforts countering an argument that the individual considers irrelevant to start with. This  will become an important problem if there are not only two but many a priori possible arguments $a_k$ for $\nprop$, a small subset of which is considered relevant by any given individual, each with their own counter-argument $a'_k$ for $\prop$.

%For a partial illustration about an ambitious setting, consider a decision to be taken about the relevance of some concrete action to help cope with climate change, and let $I$ be a set of citizens interested in the climate issue. The complete set of all arguments that will be considered relevant by at least one individual may well be very large. It may also be that, for each individual who considers relevant some arguments $A$ against the proposed concrete action, it is possible to give counter-arguments $A'$ that will convince the individual to opt for the proposal.
One of the settings discussed by \citet{thaler_nudge_2009} permits to illustrate this point more concretely.
Let $I$ be a set of workers interested in establishing a savings plan. 
The authors claim that workers tend to not save enough because of psychological biases. 
It is therefore tempting to conjecture that the deliberated choice of those workers is to save more, in the sense defined here, than what their natural choice (the one revealed in absence of arguments) leads to.
A difficulty related to this decision problem is that the complete set of all arguments that will be considered relevant by at least one individual for such a complex issue may well be very large. It may also be that, for each individual who considers relevant some arguments $A$ for saving not much (i.e., spending more in the short term), it is possible to give counter-arguments $A'$ that will convince the individual to opt for saving more.
But no single argument could be decisive for each $i \in I$: it could be that only counter-arguments built specifically according to what a given individual considers relevant could convince him, given the complexity of the issue.
This illustrates that the deliberated choice of each individuals in $I$ may be to save more even though no static theory (\cref{def:static}) may achieve showing it, as it could be that no such theory is truthful.

@@

Define $P = \set{\phi_P, \phi_D, 0}$.

Define $\ar_e$ as a text arguing that the choice between $P$ and $D$ should be determined by their expected revenue, namely, $\$3.959$ for $P$ and $\$4$ for $D$.
\footnote{Here is a possible text for $\ar_e$. “Imagine you play $P$ 100 times. On average, you would win 99 times and lose once, which means gaining 99 × \$4 = \$396 and losing 1 × \$0.10 = \$0.10, hence a total net gain of \$395.90. Thus, your average net gain for the $P$ bet would be \$3.9590. [Similar text for $D$.] Therefore, we suggest you choose $D$ for a higher expected revenue.”}
Define $\ar_u$ as a text arguing that the utility of money should matter, not the revenue, and arguing that a loss hurts more than a gain. 
\footnote{Here is a possible text for $\ar_u$. “$D$ is better than $P$ in terms of expected gain, but it is in general absurd to reason in terms of expected gain. For example, consider a choice between winning $10^6$ dollars with probability $0.8$ and winning $2 × 10^6$ dollars with probability $0.4$. These two lotteries have the same expected gain, but the first one is clearly preferrable. Therefore, we recommend to rather pick $P$ as it presents a much better chance of earning money”.}
Define a counter-argument $\ar_l$ to $\ar_u$, indicating that for small monetary amounts, it is perfectly reasonable to consider utility as linear, especially given that in this experiment, multiple lotteries will be played.
\footnote{Here is a possible text for $\ar_l$. “The lottery $D$ has a higher expected revenue, and thus, a higher utility than $P$, because for small monetary amounts, it is perfectly reasonable to consider utility as linear, especially given that in this experiment, multiple lotteries will be played.”}
Define $\ar$ so as to include the contents of $\ar_e$, $\ar_u$ and $\ar_l$.
\footnote{Here is a possible text for $\ar$. [Text of $\ar_e$] “Note that one could say that the utility of money should matter, but for small monetary amounts, it is perfectly reasonable to consider utility as linear, especially given that in this experiment, multiple lotteries will be played.”}
Define $\allargs = \set{\ar_e, \ar_u, \ar_l, \ar}$.

@@

Assume that some individuals $i \in I$ do not see the point of argument $\ar_u$ and some do but are convinced by the counter-argument $\ar_l$. Accordingly, define $\ileadsto[e]$, representing the first kind of individuals, as follows. $\forall \hist \in \allhist: \hist \ileadsto[e] D ⇔ [\ar_e \in R(\hist)]$, and $\hist \ileadsto[e] P$ otherwise. Note that in particular, if $\hist$ contains $\ar$ but not $\ar_e$, $\ar \ileadsto[e] P$, as we assume such individuals are confused by the longer argument $\ar$ referring to abstract notions and therefore opt for $P$, falling back to the common psychological bias.
The second kind of individuals is represented by $\ileadsto[u]$ defined as $\forall \hist \in \allhist: \hist \ileadsto[u] D ⇔ [\ar_u \notin R(\hist) \lor \ar \in R(\hist) \lor \set{\ar_u, \ar_l} \subseteq R(\hist)]$, and $\hist \ileadsto[u] P$ otherwise.
We assume here that such individuals think about $\ar_u$ as a correct argument in favor of $P$ until shown the rebuttal $\ar_l$ (or $\ar$, which contains a similar rebuttal).
Assuming finally $\allleadsto$ as equal to $\ileadsto[e]$ for individuals of the first kind and $\ileadsto[u]$ otherwise concludes the definition of the exposure protocol $(I, \allargs, {\allleadsto})$.

@@

This definition of truthfulness is adopted because of our goal of distinguishing empirical claims. Truthfulness, so defined, can be falsified empirically: if any of the set of claims embedded in this definition does not hold, it is possible to determine this empirically.
